---
title: "charport"
output:
  html_vignette:
    css: !expr if (identical(Sys.getenv("IN_PKGDOWN"), "true")) NULL else "charport-vignette.css"
    keep_md: no
  rmarkdown::github_document: default
vignette: >
  %\VignetteIndexEntry{charport}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, setup, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

IS_GITHUB <- nzchar(Sys.getenv("IS_GITHUB"))
IN_PKGDOWN <- identical(Sys.getenv("IN_PKGDOWN"), "true")
developer_guide <- if (IS_GITHUB) {
  "https://charport.github.io/charport/articles/developer-guide.html"
} else {
  "developer-guide.html"
}
access_diagram <- if (IS_GITHUB) {
  "man/figures/altrep-string-access.svg"
} else {
  "../man/figures/altrep-string-access.svg"
}
benchmark_figure <- if (IS_GITHUB) {
  "man/figures/bench.png"
} else {
  "../man/figures/bench.png"
}
```

```{r, echo = FALSE, results = "asis", eval = !IN_PKGDOWN}
logo <- if (IS_GITHUB) "man/figures/logo.svg" else "../man/figures/logo.svg"
cat(sprintf(
  '<img src="%s" align="right" width="160" alt="charport logo" />\n\n',
  logo
))
```

```{r, echo = FALSE, results = "asis", eval = IS_GITHUB}
cat('
[![R-CMD-check](https://github.com/charport/charport/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/charport/charport/actions)
')
```

## ALTREP string interoperability

`charport` provides a shared read interface for ALTREP character vectors.
Backend packages register their ALTREP classes, and compiled consumers read
string bytes through `charport::Reader` without depending on the backend's
storage layout. Plain and unregistered vectors fall back to R's ordinary string
representation.

*The work in this package is funded by the R Consortium Infrastructure Steering
Committee.*

## ALTREP string access

The diagram compares ordinary R access, which may materialize an ALTREP vector,
with access through a registered `charport::Reader` backend.

```{r, echo = FALSE, results = "asis"}
cat(sprintf(
  '<img src="%s" alt="ALTREP string access diagram" />\n\n',
  access_diagram
))
```

The benchmark uses the `enwik8` dataset, the first 100 million bytes of
Wikipedia split by line. The read baseline is `STRING_ELT` over an
unmaterialized `charvec` (a built-in ALTREP class). The `charport::Reader`
paths read the same data without materialization.

On the write path, writing standard R strings via `SET_STRING_ELT` is the
baseline, compared to writing the same data to `charvec`.

```{r, echo = FALSE, results = "asis"}
cat(sprintf(
  '<img src="%s" alt="charport benchmark" />\n\n',
  benchmark_figure
))
```

The results measure materialization and construction costs for this corpus and
are not a general performance guarantee.

## A broker for ALTREP strings

`charport` acts as a small broker for ALTREP strings:

1. A package registers the ALTREP string class it owns via `register_altrep`.
2. A consumer package reads string data through `charport::Reader`, without
   knowing that class's storage layout.
3. Registered ALTREP classes can be read directly; ordinary vectors and
   unregistered ALTREP classes fall back to standard R behavior.

The contract covers bytes, encoding marks, pointer lifetime, and access
capabilities. `charport` does not define string semantics, locale policy,
normalization, or a new user-facing string API.

## `charvec`: a reference ALTREP character vector

`charport` also includes `charvec`, which stores string data in native blocks
alongside a vector of metadata.

To R, a `charvec` behaves like an ordinary character vector. In compiled code,
a `charvec` can be constructed serially or across multiple worker threads, then
read through `charport::Reader` without materialization. It serves as both a
reference implementation and an output type for packages that produce string
data.

## For package developers

Package authors can use `charport` from either side of the interface.

ALTREP string classes register through `register_altrep`. Consumer packages
read through `charport::Reader`, and packages that construct string output can
use `charvec` directly.

The package developer guide covers registration, reading, fallback behavior,
pointer lifetime, thread safety, and the `charvec` builder.

```{r, echo = FALSE, results = "asis"}
cat(sprintf(
  "- [Package developer guide](%s), also listed by `utils::vignette(package = \"charport\")`.\n",
  developer_guide
))
```
