---
title: "charport package developer guide"
output:
  html_vignette:
    css: !expr if (identical(Sys.getenv("IN_PKGDOWN"), "true")) NULL else "charport-vignette.css"
    tabset: true
vignette: >
  %\VignetteIndexEntry{charport package developer guide}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

developer_flow <- "../man/figures/developer-flow.svg"
```

This guide is for packages that read ALTREP character vectors, register an
ALTREP string class, or build `charvec` output.

`charport` defines an interface between ALTREP string producers and consumers.
Producers register a way to view their string data, and consumers read those
views without depending on the producer's storage layout or forcing ordinary R
string materialization.

A package may produce an ALTREP string class, consume ALTREP strings, or use
`charvec`, a built-in ALTREP class, as its own output representation. The
diagram below shows how those roles meet through charport.

```{r, echo = FALSE, results = "asis"}
cat(sprintf(
  '<img src="%s" alt="charport developer integration flowchart" />\n\n',
  developer_flow
))
```

The guide covers three use cases:

- [Using `Reader`](#universal-reader): read ALTREP
  strings through one C++ interface.
- [Registering an ALTREP class](#altrep-registration): make an ALTREP string
  class readable by other packages through `charport`.
- [Using `charvec` directly](#charvec-direct): build the included `charvec`
  class with serial or multithreaded builders.

All three use cases start the same way: link to `charport`, include the
umbrella header, and check the ABI once when your package loads.

## Linking to `charport` {#linking-to-charport}

Use the C++ interface in `charport.h`.

In `DESCRIPTION`, declare both the runtime dependency and the header
dependency:

```text
Imports:
    charport
LinkingTo:
    charport
```

In `NAMESPACE`, import `charport` so its runtime entry points are available:

### Namespace setup {.tabset}

#### roxygen2
```r
#' @useDynLib yourpkg, .registration = TRUE
#' @import charport
"_PACKAGE"
```

#### NAMESPACE
```text
useDynLib(yourpkg, .registration = TRUE)
import(charport)
```

### ABI check hook {.tabset}

`check_abi()` compares the ABI version compiled into your package with the ABI
provided by the installed `charport`. It returns `true` when they match and
`false` when they do not. Your package decides how to report a mismatch.

#### Manual `.onLoad`

```cpp
#include "charport.h"

extern "C" SEXP C_check_charport_abi(void) {
  return Rf_ScalarLogical(charport::check_abi() ? TRUE : FALSE);
}
```

Register `C_check_charport_abi` in the package's `.Call` table, then add:

```r
.onLoad <- function(libname, pkgname) {
  if (!isTRUE(.Call(C_check_charport_abi))) {
    stop(
      "yourpkg was built against an incompatible charport ABI",
      call. = FALSE
    )
  }
}
```

#### Rcpp init

```cpp
#include "charport.h"

// [[Rcpp::init]]
void init_yourpkg(DllInfo * dll) {
  if(!charport::check_abi()) {
    Rf_error("yourpkg was built against an incompatible charport ABI");
  }
  (void)dll;
}
```

#### cpp11 init

```cpp
#include "charport.h"

[[cpp11::init]]
void init_yourpkg(DllInfo * dll) {
  if(!charport::check_abi()) {
    Rf_error("yourpkg was built against an incompatible charport ABI");
  }
  (void)dll;
}
```

#### Manual init

```cpp
#include "charport.h"

extern "C" void R_init_yourpkg(DllInfo * dll) {
  R_registerRoutines(dll, NULL, call_entries, NULL, NULL);
  R_useDynamicSymbols(dll, FALSE);
  if(!charport::check_abi()) {
    Rf_error("yourpkg was built against an incompatible charport ABI");
  }
  R_forceSymbols(dll, TRUE);
}
```

### ABI version

The current ABI version is `CHARPORT_ABI_VERSION=1`. Any layout or contract
break bumps this value. `check_abi()` returns `false` when the version compiled
into `yourpkg` does not match the installed `charport` version.

## Using `Reader` {#universal-reader}

`charport::Reader` reads string data from a registered ALTREP class or an
ordinary R character vector. Construct it from a character vector, then access
elements by iteration or index.

```cpp
using namespace charport;
Reader r(robj);
for(StrView s : r) {
  if(!s.is_na()) {
    // s.ptr, s.len, s.enc
  }
}

for(R_xlen_t i=0; i<r.size(); ++i) {
  StrView s = r[i];
  // ...
}
```

For scalar inputs, `Reader::read_scalar()` reads the first element and returns a
`StrView`:

```cpp
StrView s = Reader::read_scalar(x);
```

`read_scalar()` expects a non-empty character vector and should be called on
the main R thread. It uses the registered ALTREP path when it can return a
persistent view without keeping reader state alive; otherwise it uses ordinary
R string access.

For a registered ALTREP class, `Reader` gives you non-materializing views. For
a plain vector, a materialized ALTREP or an ALTREP that is not registered,
`Reader` falls back to ordinary access via `STRING_PTR_RO`.

`Reader` owns the active borrow and is move-only. Most code should keep it as a
local stack object and let its destructor clean up any state supplied by the
registered class.

### Reader properties

`Reader::kind()` reports how the current `Reader` was resolved. It describes
the read path chosen when the `Reader` was constructed.

| Value | Meaning |
|---|---|
| `charport::ReaderKind::PLAIN` | Ordinary non-ALTREP character vector. |
| `charport::ReaderKind::MATERIALIZED_ALTREP` | ALTREP character vector whose ordinary string storage is already available. |
| `charport::ReaderKind::REGISTERED_ALTREP` | ALTREP character vector served by a `charport`-registered class. |
| `charport::ReaderKind::FALLBACK_ALTREP` | ALTREP character vector not served directly; resolving the `Reader` falls back to ordinary R string access. |

`Reader::view_persistence()` is true when returned `StrView` pointers remain
valid for the lifetime of the `Reader`.

`Reader::thread_safe_access()` is true when `Reader` access may happen from
outside the main thread and concurrently from multiple threads.

`Reader::reentrant()` is a combined convenience check when `view_persistence()`
and `thread_safe_access()` are true.

### `StrView`

`charport::StrView` is the type returned by `Reader` element access. It is a
non-owning view of bytes owned by the character vector.

Each `StrView` contains:

1. `ptr`, a pointer to the string data
2. `len`, the string length in bytes
3. `enc`, the encoding mark for those bytes

`ptr` is not guaranteed to be null-terminated; use `len` for its byte length.

`enc` is an enum similar to, but not identical to, base R's `cetype_t`. The
following values are defined:

| Value | Meaning |
|---|---|
| `charport_enc::CE_NATIVE` | Native encoding based on machine locale. |
| `charport_enc::CE_UTF8` | UTF-8. |
| `charport_enc::CE_LATIN1` | Latin-1. |
| `charport_enc::CE_BYTES` | Bytes encoding. |
| `charport_enc::CE_ASCII_OR_UTF8` | ASCII/UTF-8-compatible bytes. |
| `charport_enc::CE_ASCII` | ASCII bytes, equivalent to `Rf_charIsASCII(x)`. |
| `charport_enc::CE_NA` | Missing string. `ptr` is `NULL` and `len` is zero. |

These fields carry the byte data and encoding information represented by a base
R `CHARSXP`.

### Reader lifetime and multithreading

In Rust terms, a `Reader` acts like a strict borrow. It is a temporary view of
the vector's current string storage, valid only while that storage remains
protected and unmodified. While the `Reader` is active, do not access or modify
the original R object. Calls such as `STRING_ELT`, `STRING_PTR_RO`, and
`DATAPTR_RO` can invalidate the borrow.

When `view_persistence()` is true, `StrView` pointers returned from the
`Reader` remain valid after later `Reader` access, until the active borrow
ends. This allows code to collect views on the main thread and process on
worker threads.

When `thread_safe_access()` is true, `Reader` access itself may happen off the
main thread and concurrently. `reentrant()` returns true only when both
`view_persistence()` and `thread_safe_access()` are true.

The fallback `Reader` path has persistent views but is main-thread-only.

## Registering an ALTREP class {#altrep-registration}

A package can register its ALTREP class to make non-materializing access
available to `charport::Reader`. Registration can happen directly in native
package initialization, or from `.onLoad()` via a small native routine.

The registration function has the following signature:

```cpp
void charport::register_altrep(
  R_altrep_class_t cls,
  charport_reader_init_fn reader_init,       // (SEXP x) -> void*
  charport_reader_get_fn reader_get,         // (void* state, R_xlen_t i) -> StrView
  charport_reader_release_fn reader_release, // (void* state) -> void
  bool view_persistence,
  bool thread_safe_access
);
```

`reader_init(SEXP x)` starts the borrow and returns the state that
`reader_get()` needs. When `reader_release` is `NULL`, the state is borrowed
and must remain owned by the ALTREP object. When `reader_release` is
non-`NULL`, the state is a per-`Reader` native context and `charport::Reader`
releases it when the borrow ends. Returning `NULL` from `reader_init` means
this particular vector cannot satisfy the reader contract, so `charport` falls
back to default access using `STRING_PTR_RO`.

`reader_get(void * state, R_xlen_t i)` returns the view for element `i`. It
receives only reader state and the index.

`reader_release(void * state)` is optional. Use it when `reader_init` allocates
a native reader context, for example to hold temporary decoded bytes.

If `cls` is already registered, `register_altrep()` throws an error. Duplicate
registration probably means the package registered from more than one load hook
or failed to unregister before reload.

### Registration location {.tabset}

#### .onLoad hook registration example

An `.onLoad` hook can call a small native registration routine:

```cpp
extern "C" SEXP C_register_yourpkg_altrep(void) {
  if(!charport::check_abi()) {
    Rf_error("yourpkg was built against an incompatible charport ABI");
  }
  charport::register_altrep(
    your_altrep, reader_init, reader_get, nullptr, true, true
  );
  return R_NilValue;
}

extern "C" SEXP C_unregister_yourpkg_altrep(void) {
  charport::unregister_altrep(your_altrep);
  return R_NilValue;
}
```

```r
.onLoad <- function(libname, pkgname) {
  .Call(C_register_yourpkg_altrep)
}

.onUnload <- function(libpath) {
  .Call(C_unregister_yourpkg_altrep)
}
```

#### Init hook registration example

Native package initialization is the usual registration point.

```cpp
static R_altrep_class_t your_altrep;

extern "C" void R_init_yourpkg(DllInfo * dll) {
  // Other init hook operations
  // ...
  your_altrep = R_make_altstring_class("yourclass", "yourpkg", dll);
  if(!charport::check_abi()) {
    Rf_error("yourpkg was built against an incompatible charport ABI");
  }
  charport::register_altrep(
    your_altrep, reader_init, reader_get, nullptr, true, true
  );
}

extern "C" void R_unload_yourpkg(DllInfo * dll) {
  charport::unregister_altrep(your_altrep);
  (void)dll;
}
```

Rcpp and cpp11 initialization hooks can perform the same registration.

### Error and unwind rules

Every function pointer registered with `charport`, and every C++ function
called through `.Call`, is entered through a C interface. If that code can
throw a C++ exception, catch it before returning through the C interface.
Convert the failure to an R error, a `NULL` init result, or another explicit
status before the registered function or `.Call` entry point returns.

`reader_init(SEXP)` runs on the main R thread and may report failure with
`Rf_error()`. `reader_get(state, i)` should be total for valid `i`; it must not
call the R API and must not throw C++ exceptions. `reader_release(state)`
should also be R-free and exception-free.

## Using `charvec` directly {#charvec-direct}

`charvec` is `charport`'s included ALTREP string class. Packages can use it as
an output representation without implementing another ALTREP backend.

Internally, a `charvec` object owns two things:

1. One or more byte slices that hold string data.
2. A fixed array of string records.

The shape is roughly:

```cpp
struct charvec_data {
  char * slice_head;     // linked list of owned byte blocks
  StrView records[n];    // one record per vector element
};
```

For non-empty strings, `records[i].ptr` points into one of the owned byte
slices. Empty strings point at shared empty storage. Missing strings are stored
as `{nullptr, 0, charport_enc::CE_NA}`.

Starting at `slice_head`, each slice is one allocation: a pointer storing the
next-slice address, followed by the string data bytes. When the active slice
does not have enough room, `charvec` allocates a new slice and links it at the
head. For multi-string vectors, the initial slice size scales with vector
length and is capped at 16 KiB. Later slices grow to a size of half the bytes
already allocated, rounded to 64-byte boundaries, with a 256 KiB per-slice cap.

For one-element output, `charvec::build_scalar()` is the smallest helper:

```cpp
SEXP out = charvec::build_scalar(view);
```

Like the builder `to_sexp()` methods, call it on the main R thread.

### Constructing a `charvec` with `charvec::Builder`

`charvec::Builder` builds a `charvec` serially, one string at a time. The
constructor takes the vector length and creates an initial slice. Insert a
string with either `set()` or `reserve()`.

Use `set()` to copy an existing buffer.

```cpp
charvec::Builder b(n);
b.set(i, ptr, len, charport_enc::CE_UTF8);
```

Use `reserve()` when the output bytes can be written directly into the store:

```cpp
char * dst = b.reserve(i, len, charport_enc::CE_UTF8);
// write to dst directly
```

`to_sexp()` returns the ALTREP R vector. Call `reset()` to reuse the builder.

```cpp
SEXP out = b.to_sexp();
b.reset(n);
```

### Constructing a `charvec` with `charvec::DirectBuilder`

Use `charvec::DirectBuilder` when the total byte size is known in advance or
the caller needs to control slice growth. It exposes slice pointers and the
record array directly.

```cpp
charvec::DirectBuilder b(n, total_bytes);

char * data = b.data_slice(0); // 0 is the index of the slice
StrView * records = b.records();

std::memcpy(data, src, total_bytes);
records[0] = {data + offset0, len0, charport_enc::CE_UTF8};
records[1] = {nullptr, 0, charport_enc::CE_NA};

SEXP out = b.to_sexp();
b.reset(n, total_bytes);
```

`allocate_next_slice(bytes)` allocates another owned slice and returns its data
pointer.

### Constructing a `charvec` with `charvec::ParallelBuilder`

Use `charvec::ParallelBuilder` for parallel construction. The builder owns a
fixed set of shards. Each worker uses a distinct shard index with `set()` or
`reserve()`. `to_sexp()` combines the shards on the main thread.

```cpp
charvec::ParallelBuilder b(n, n_workers);

// worker_id is 0 .. n_workers-1
parallel_for_each_worker(worker_id, n_workers) {
  R_xlen_t begin = n * worker_id / n_workers;
  R_xlen_t end = n * (worker_id + 1) / n_workers;

  for(R_xlen_t i = begin; i < end; ++i) {
    char * dst = b.reserve(worker_id, i, len[i], charport_enc::CE_UTF8);
    // write this worker's bytes into dst
  }
}

SEXP out = b.to_sexp();
```

## Summary checklists

### Using `Reader`

1. Check `charport::check_abi()` at load time and report a mismatch before
   using the `charport` interface.
2. Construct `Reader` objects on the main R thread.
3. Use `Reader::read_scalar()` when you only need the first element.
4. Respect the borrow rule: no R access to the source vector or aliases during
   active reads.
5. Call `Reader` access from worker threads only when
   `r.thread_safe_access()` is true.

### Registering an ALTREP class

1. Register in `R_init_yourpkg()` and unregister in `R_unload_yourpkg()`.
2. Return `NULL` from `reader_init()` when a specific vector instance cannot
   satisfy the `Reader` contract.
3. Make `get` and `release` free of R API and exceptions for valid indices.
4. Set `view_persistence = true` when returned views remain valid until
   the borrow ends.
5. Set `thread_safe_access = true` when concurrent calls are safe.

### Using `charvec` directly

1. Use `charvec::build_scalar()` for one-element output.
2. Use `charvec::Builder` for ordinary serial construction.
3. Use `charvec::DirectBuilder` when you want to manage slices and records
   directly.
4. Use `charvec::ParallelBuilder` for parallel construction.
5. Store missing values as `{nullptr, 0, charport_enc::CE_NA}` and set encoding
   marks deliberately.
6. Call `to_sexp()` only after all writes are complete, and on the main R
   thread.
