---
title: "charport package developer guide"
output:
  html_vignette:
    css: !expr if (identical(Sys.getenv("IN_PKGDOWN"), "true")) NULL else "charport-vignette.css"
    tabset: true
vignette: >
  %\VignetteIndexEntry{charport package developer guide}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

developer_flow <- "../man/figures/developer-flow.svg"
```

This guide is for packages that read ALTREP character vectors, register an
ALTREP string class, or build `charvec` output.

`charport` defines an interface between ALTREP string producers and consumers.
Producers register a way to view their string data, and consumers read those
views without depending on the producer's storage layout or forcing ordinary R
string materialization.

For the design decisions behind that contract, see [Design
Rationale](design-rationale.html).

A package may produce an ALTREP string class, consume ALTREP strings, or use
`charvec`, a built-in ALTREP class, as its own output representation. The
diagram below shows how those roles meet through charport.

```{r, echo = FALSE, results = "asis"}
cat(sprintf(
  '<img src="%s" alt="charport developer integration flowchart" />\n\n',
  developer_flow
))
```

The guide covers three use cases:

- [Using `Reader`](#universal-reader): read ALTREP
  strings through one C++ interface.
- [Registering an ALTREP class](#altrep-registration): make an ALTREP string
  class readable by other packages through `charport`.
- [Using `charvec` directly](#charvec-direct): build the included `charvec`
  class with serial or multithreaded builders.

All three use cases start the same way: link to `charport`, include the
umbrella header, and check the ABI once when your package loads.

## Linking to `charport` {#linking-to-charport}

Use the C++ interface in `charport.h`.

In `DESCRIPTION`, declare both the runtime dependency and the header
dependency:

```text
Imports:
    charport
LinkingTo:
    charport
```

In `NAMESPACE`, import `charport` so its runtime entry points are available:

### Namespace setup {.tabset}

#### roxygen2
```r
#' @useDynLib yourpkg, .registration = TRUE
#' @import charport
"_PACKAGE"
```

#### NAMESPACE
```text
useDynLib(yourpkg, .registration = TRUE)
import(charport)
```

### ABI check hook {.tabset}

`check_abi()` compares the ABI version compiled into your package with the ABI
provided by the installed `charport`. It returns `true` when they match and
`false` when they do not. Your package decides how to report a mismatch.

#### Manual `.onLoad`

```cpp
#include "charport.h"

extern "C" SEXP C_check_charport_abi(void) {
  return Rf_ScalarLogical(charport::check_abi() ? TRUE : FALSE);
}
```

Register `C_check_charport_abi` in the package's `.Call` table, then add:

```r
.onLoad <- function(libname, pkgname) {
  if (!isTRUE(.Call(C_check_charport_abi))) {
    stop(
      "yourpkg was built against an incompatible charport ABI",
      call. = FALSE
    )
  }
}
```

#### Rcpp init

```cpp
#include "charport.h"

// [[Rcpp::init]]
void init_yourpkg(DllInfo * dll) {
  if(!charport::check_abi()) {
    Rf_error("yourpkg was built against an incompatible charport ABI");
  }
  (void)dll;
}
```

#### cpp11 init

```cpp
#include "charport.h"

[[cpp11::init]]
void init_yourpkg(DllInfo * dll) {
  if(!charport::check_abi()) {
    Rf_error("yourpkg was built against an incompatible charport ABI");
  }
  (void)dll;
}
```

#### Manual init

```cpp
#include "charport.h"

extern "C" void R_init_yourpkg(DllInfo * dll) {
  R_registerRoutines(dll, NULL, call_entries, NULL, NULL);
  R_useDynamicSymbols(dll, FALSE);
  if(!charport::check_abi()) {
    Rf_error("yourpkg was built against an incompatible charport ABI");
  }
  R_forceSymbols(dll, TRUE);
}
```

### ABI version

The current ABI version is `CHARPORT_ABI_VERSION=1`. Any layout or contract
break bumps this value. `check_abi()` returns `false` when the version compiled
into `yourpkg` does not match the installed `charport` version.

## Using `Reader` {#universal-reader}

`charport::Reader` reads string data from a registered ALTREP class or an
ordinary R character vector.

The main access pattern is block access. Construct a `Reader`, choose a block
size that works for your algorithm, and ask `Reader` to fill caller-owned
buffers.

```cpp
using namespace charport;

Reader r(robj);
const R_xlen_t block_size = 4096;
ByteViews byteviews(block_size);

for(R_xlen_t start = 0; start < r.size(); start += block_size) {
  R_xlen_t n = std::min(block_size, r.size() - start);
  r.byteviews(start, n, byteviews);

  for(R_xlen_t j = 0; j < n; ++j) {
    ByteView b = byteviews[j];
    if(!b.is_na()) {
      // b.ptr, b.len
    }
  }
}
```

Use the narrowest access that fits the work:

- `byteviews()` when you need bytes and lengths.
- `lengths()` when you only need element sizes and missingness.
- `encodings()` when you only need encoding marks.

There are two block forms. Range access reads a contiguous 0-based interval:

```cpp
r.views(start, size, strviews);
r.byteviews(start, size, byteviews);
r.lengths(start, size, out_len);
r.encodings(start, size, out_enc);
```

Indexed access gathers 0-based indices:

```cpp
R_xlen_t idx[] = {0, 10, 20};
r.byteviews(idx, 3, byteviews);
```

Scalar methods are convenience wrappers for code that really is element-wise:

```cpp
for(R_xlen_t i = 0; i < r.size(); ++i) {
  StrView s = r.view(i);
  if(!s.is_na()) {
    // s.ptr, s.len, s.enc
  }
}

for(R_xlen_t i = 0; i < r.size(); ++i) {
  ByteView b = r.byteview(i);
  if(!b.is_na()) {
    // b.ptr, b.len
  }
}

for(R_xlen_t i = 0; i < r.size(); ++i) {
  int len = r.length(i);
  if(len >= 0) {
    // len is the element length in bytes
  }
}

for(R_xlen_t i = 0; i < r.size(); ++i) {
  cetype_ext_t enc = r.encoding(i);
  // enc is cetype_ext_t::CE_NA for missing strings
}
```

`operator[]` and range-for iteration are aliases for the full `StrView`
interface. They are handy, but block access is usually the better default for
large vectors.

For a registered ALTREP class, `Reader` gives you non-materializing views. For
a plain vector, a materialized ALTREP or an ALTREP that is not registered,
`Reader` falls back to ordinary access via `STRING_PTR_RO`.

`Reader` owns the active borrow and is move-only. Most code should keep it as a
local stack object and let its destructor clean up any state supplied by the
registered class.

### Reader information

Most code should not branch on the source vector's implementation. Construct a
`Reader` and use the guarantees it reports.

`charport::sexp_info(x)` is available for diagnostics and preflight checks when
you need to understand a source vector without forcing it:

```cpp
charport::SexpInfo info = charport::sexp_info(x);
```

It reports whether `x` is a character vector, whether it is ALTREP, whether
ordinary string pointer storage is already available, whether a registered
ALTREP class matches, the reader capability flags, and whether the registered
reader owns per-`Reader` state. ALTREP class and package names are reported as
borrowed `const char *` pointers, or `NULL` when unavailable.

`sexp_info()` is non-forcing. It does not call `STRING_ELT()`,
`STRING_PTR_RO()`, `DATAPTR()`, `reader_init()`, or the reader callbacks.
`is_materialized` means `DATAPTR_OR_NULL(x) != NULL`: ordinary string pointer
storage is available without forcing. It does not mean that no element has ever
been lazily cached by an ALTREP class.

`Reader::persistent_views()` is true when returned byte pointers remain valid
for the lifetime of the `Reader`, even after later `Reader` access calls.

`Reader::concurrent_access()` is true when `Reader` access calls may happen
from worker threads and from multiple threads at the same time.

`Reader::reentrant()` is a convenience check that returns true when both
capabilities are true.

### `StrView`

`charport::StrView` is the type returned by `Reader::view(i)`,
`Reader::operator[]`, and range-for iteration. It is a non-owning view of bytes
owned by the character vector.

Each `StrView` contains:

1. `ptr`, a pointer to the string data
2. `len`, the string length in bytes
3. `enc`, the encoding mark for those bytes

`ptr` is not guaranteed to be null-terminated; use `len` for its byte length.

`enc` is an enum similar to, but not identical to, base R's `cetype_t`. The
following values are defined:

| Value | Meaning |
|---|---|
| `cetype_ext_t::CE_NATIVE` | Native encoding based on machine locale. |
| `cetype_ext_t::CE_UTF8` | UTF-8. |
| `cetype_ext_t::CE_LATIN1` | Latin-1. |
| `cetype_ext_t::CE_BYTES` | Bytes encoding. |
| `cetype_ext_t::CE_ASCII_OR_UTF8` | ASCII/UTF-8-compatible bytes. |
| `cetype_ext_t::CE_ASCII` | ASCII bytes, equivalent to `Rf_charIsASCII(x)`. |
| `cetype_ext_t::CE_NA` | Missing string. `ptr` is `NULL` and `len` is `NA_INTEGER`. |

These fields carry the byte data and encoding information represented by a base
R `CHARSXP`.

### Byte views, lengths, and encodings

`charport::ByteView` is the type returned by `Reader::byteview(i)` and by
`ByteViews::operator[]`. It contains only `ptr` and `len`. Like `StrView`,
`ptr` is not guaranteed to be null-terminated, so `len` is required for every
byte operation.

`Reader::length(i)` returns the element length in bytes without returning a
pointer. Missing strings return `NA_INTEGER`.

`Reader::encoding(i)` returns the element encoding mark without returning a
pointer. Missing strings return `cetype_ext_t::CE_NA`.

### Reader lifetime and multithreading

In Rust terms, a `Reader` acts like a strict borrow. It is a temporary view of
the vector's current string storage, valid only while that storage remains
protected and unmodified. While the `Reader` is active, do not access or modify
the original R object. Calls such as `STRING_ELT`, `STRING_PTR_RO`, and
`DATAPTR_RO` can invalidate the borrow.

When `persistent_views()` is true, byte pointers returned from the `Reader`
remain valid after later `Reader` access, until the active borrow ends. This
allows code to collect views on the main thread and process on worker threads.

When `concurrent_access()` is true, `Reader` access itself may happen off the
main thread and concurrently. `reentrant()` returns true only when both
`persistent_views()` and `concurrent_access()` are true.

The fallback `Reader` path has persistent views but is main-thread-only.

Attributes are orthogonal to the `Reader` borrow. You can inspect attributes on
the original `SEXP` while a `Reader` exists, as long as you do not force or
modify the character data itself.

## Registering an ALTREP class {#altrep-registration}

A package can register its ALTREP class to make non-materializing access
available to `charport::Reader`. Registration can happen directly in native
package initialization, or from `.onLoad()` via a small native routine.

The registration function has the following signature:

```cpp
void charport::register_altrep(
  R_altrep_class_t cls,
  charport_reader_state_fns state_fns,
  charport_reader_access_fns access_fns,
  charport_reader_capabilities capabilities
);
```

The grouped structs are plain C structs:

```cpp
struct charport_reader_state_fns {
  charport_reader_init_fn init;
  charport_reader_release_fn release;
};

struct charport_reader_access_fns {
  charport_reader_range_fn range;
  charport_reader_index_fn index;
};

struct charport_reader_capabilities {
  bool persistent_views;
  bool concurrent_access;
};
```

`init(SEXP x)` starts the borrow and returns the state that the reader
callbacks need. When `release` is `NULL`, the state is borrowed and must remain
owned by the ALTREP object. When `release` is non-`NULL`, the state is a
per-`Reader` native context and `charport::Reader` releases it when the borrow
ends.

Returning `NULL` from `init` means this particular vector cannot satisfy the
reader contract, so `charport` falls back to default access using
`STRING_PTR_RO`.

Range callbacks fill caller-owned output arrays for the 0-based interval
`[start, start + size)`. Indexed callbacks fill caller-owned output arrays for
the supplied 0-based indices:

```cpp
void range(void * state, R_xlen_t start, R_xlen_t size,
           const char ** out_ptrs,
           int * out_lens,
           cetype_ext_t * out_encs);

void index(void * state, const R_xlen_t * indices, R_xlen_t size,
           const char ** out_ptrs,
           int * out_lens,
           cetype_ext_t * out_encs);
```

Any output pointer may be `NULL`; only fill arrays that were requested. Missing
strings are reported as `ptr = nullptr`, `len = NA_INTEGER`, and `enc =
cetype_ext_t::CE_NA`.

Scalar `Reader` methods are convenience wrappers over a range request of size
one. Registered classes do not provide separate scalar callbacks.

If `cls` is already registered, `register_altrep()` throws an error. Duplicate
registration probably means the package registered from more than one load hook
or failed to unregister before reload.

### Registration location {.tabset}

#### .onLoad hook registration example

An `.onLoad` hook can call a small native registration routine:

```cpp
extern "C" SEXP C_register_yourpkg_altrep(void) {
  if(!charport::check_abi()) {
    Rf_error("yourpkg was built against an incompatible charport ABI");
  }
  charport::register_altrep(
    your_altrep,
    charport_reader_state_fns{reader_init, nullptr},
    charport_reader_access_fns{reader_range, reader_index},
    charport_reader_capabilities{true, true}
  );
  return R_NilValue;
}

extern "C" SEXP C_unregister_yourpkg_altrep(void) {
  charport::unregister_altrep(your_altrep);
  return R_NilValue;
}
```

```r
.onLoad <- function(libname, pkgname) {
  .Call(C_register_yourpkg_altrep)
}

.onUnload <- function(libpath) {
  .Call(C_unregister_yourpkg_altrep)
}
```

#### Init hook registration example

Native package initialization is the usual registration point.

```cpp
static R_altrep_class_t your_altrep;

extern "C" void R_init_yourpkg(DllInfo * dll) {
  // Other init hook operations
  // ...
  your_altrep = R_make_altstring_class("yourclass", "yourpkg", dll);
  if(!charport::check_abi()) {
    Rf_error("yourpkg was built against an incompatible charport ABI");
  }
  charport::register_altrep(
    your_altrep,
    charport_reader_state_fns{reader_init, nullptr},
    charport_reader_access_fns{reader_range, reader_index},
    charport_reader_capabilities{true, true}
  );
}

extern "C" void R_unload_yourpkg(DllInfo * dll) {
  charport::unregister_altrep(your_altrep);
  (void)dll;
}
```

Rcpp and cpp11 initialization hooks can perform the same registration.

### Error and unwind rules

Every function pointer registered with `charport`, and every C++ function
called through `.Call`, is entered through a C interface. If that code can
throw a C++ exception, catch it before returning through the C interface.
Convert the failure to an R error, a `NULL` init result, or another explicit
status before the registered function or `.Call` entry point returns.

`init(SEXP)` runs on the main R thread and may report failure with
`Rf_error()`. Range and indexed reader callbacks should be total for valid
indices; they must not call the R API and must not throw C++ exceptions.
`release(state)` should also be R-free and exception-free.

## Using `charvec` directly {#charvec-direct}

`charvec` is `charport`'s included ALTREP string class. Packages can use it as
an output representation without implementing another ALTREP backend.

Internally, a `charvec` object owns two things:

1. One or more byte slices that hold string data.
2. Three fixed metadata arrays: pointers, lengths, and encoding marks.

The shape is roughly:

```cpp
struct charvec_data {
  char * slice_head;       // linked list of owned byte blocks
  const char ** ptrs;      // one pointer per element
  int * lens;              // one byte length per element
  cetype_ext_t * encs;     // one encoding mark per element
};
```

For non-empty strings, `ptrs[i]` points into one of the owned byte slices.
Empty strings point at shared empty storage. Missing strings are stored as
`ptrs[i] = nullptr`, `lens[i] = NA_INTEGER`, and `encs[i] =
cetype_ext_t::CE_NA`.

Starting at `slice_head`, each slice is one allocation: a pointer storing the
next-slice address, followed by the string data bytes. When the active slice
does not have enough room, `charvec` allocates a new slice and links it at the
head. For multi-string vectors, the initial slice size scales with vector
length and is capped at 16 KiB. Later slices grow to a size of half the bytes
already allocated, rounded to 64-byte boundaries, with a 256 KiB per-slice cap.

For one-element output, `charvec::build_scalar()` is the smallest helper:

```cpp
SEXP out = charvec::build_scalar(view);
```

Like the builder `to_sexp()` methods, call it on the main R thread.

### Constructing a `charvec` with `charvec::Builder`

`charvec::Builder` builds a `charvec` serially, one string at a time. The
constructor takes the vector length; byte slices are allocated as strings are
written. Insert a string with either `set()` or `reserve()`.

Use `set()` to copy an existing buffer.

```cpp
charvec::Builder b(n);
b.set(i, ptr, len, cetype_ext_t::CE_UTF8);
```

Use `reserve()` when the output bytes can be written directly into the store:

```cpp
char * dst = b.reserve(i, len, cetype_ext_t::CE_UTF8);
// write to dst directly
```

`to_sexp()` returns the ALTREP R vector. Call `reset()` to reuse the builder.

```cpp
SEXP out = b.to_sexp();
b.reset(n);
```

### Constructing a `charvec` with `charvec::DirectBuilder`

Use `charvec::DirectBuilder` when the total byte size is known in advance or
the caller needs to control slice growth. It exposes slice data and metadata
arrays directly.

```cpp
charvec::DirectBuilder b(n, total_bytes);

char * data = b.data_slice(0); // 0 is the index of the slice
const char ** ptrs = b.ptrs();
int * lens = b.lengths();
cetype_ext_t * encs = b.encodings();

std::memcpy(data, src, total_bytes);
ptrs[0] = data + offset0;
lens[0] = len0;
encs[0] = cetype_ext_t::CE_UTF8;

ptrs[1] = nullptr;
lens[1] = NA_INTEGER;
encs[1] = cetype_ext_t::CE_NA;

SEXP out = b.to_sexp();
b.reset(n, total_bytes);
```

`allocate_next_slice(bytes)` allocates another owned slice and returns its data
pointer.

### Constructing a `charvec` with `charvec::ParallelBuilder`

Use `charvec::ParallelBuilder` for parallel construction. The builder owns a
fixed set of shards. Each worker uses a distinct shard index with `set()` or
`reserve()`. `to_sexp()` combines the shards on the main thread.

```cpp
charvec::ParallelBuilder b(n, n_workers);

// worker_id is 0 .. n_workers-1
parallel_for_each_worker(worker_id, n_workers) {
  R_xlen_t begin = n * worker_id / n_workers;
  R_xlen_t end = n * (worker_id + 1) / n_workers;

  for(R_xlen_t i = begin; i < end; ++i) {
    char * dst = b.reserve(worker_id, i, len[i], cetype_ext_t::CE_UTF8);
    // write this worker's bytes into dst
  }
}

SEXP out = b.to_sexp();
```

## Summary checklists

### Using `Reader`

1. Check `charport::check_abi()` at load time and report a mismatch before
   using the `charport` interface.
2. Construct `Reader` objects on the main R thread.
3. Prefer block access for large vectors.
4. Use scalar access only when the algorithm is naturally element-wise.
5. Respect the borrow rule: no R access to the source vector or aliases during
   active reads.
6. Call `Reader` access from worker threads only when
   `r.concurrent_access()` is true.

### Registering an ALTREP class

1. Register in `R_init_yourpkg()` and unregister in `R_unload_yourpkg()`.
2. Return `NULL` from `init()` when a specific vector instance cannot satisfy
   the `Reader` contract.
3. Make range, indexed, and release callbacks free of R API and exceptions for
   valid indices.
4. Set `persistent_views = true` when returned views remain valid until
   the borrow ends.
5. Set `concurrent_access = true` when concurrent access calls are safe.

### Using `charvec` directly

1. Use `charvec::build_scalar()` for one-element output.
2. Use `charvec::Builder` for ordinary serial construction.
3. Use `charvec::DirectBuilder` when you want to manage slices and metadata
   directly.
4. Use `charvec::ParallelBuilder` for parallel construction.
5. Store missing values as `ptr = nullptr`, `len = NA_INTEGER`, and
   `enc = cetype_ext_t::CE_NA`.
6. Call `to_sexp()` only after all writes are complete, and on the main R
   thread.
